---
title: "ReprodResearch1"
output: 
  html_document:
    keep_md: true
---
This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

## Loading and preprocessing the data
The data is loaded and tidied, converting the *interval* field from a varying 1 to 4 digit number representing a mixture hours/minutes into a simpler numeric factor which can be combined with the *date* to create a proper *time* column.

```{r}
    #load and tidy data
    myData <- read.csv("activity.csv")
    myData$interval <- as.numeric(as.factor(myData$interval))
    myData$time <- strptime(myData$date,"%Y-%m-%d") + 300 * myData$interval

```

Some snapshots of the data illustrate it's current form, including a number of missing values:

```{r echo=FALSE}
    head(myData, 5)

    tail(myData, 5)
```

In preparation for graphing the lattice library is loaded:
```{r}
    #Ensure required libraries are loaded.
    require(lattice)
```

## What is mean total number of steps taken per day?
The histogram below is prepared from the tidy data set using the following code.  It also displays the mean and median: 

```{r}    
    #summarise data by day
    mySteps<-aggregate(steps ~ date, myData, FUN = mean, rm.na = TRUE)
    hist(mySteps$steps, main = "Histogram of Steps per day", xlab = "")  
    text(70, 18, paste("Mean = ", round(mean(mySteps$steps), 2), "\n Median = ", 
                     round(median(mySteps$steps), 2)))
```

## What is the average daily activity pattern?
The data set can also be used to provide a graph of average daily activity.
```{r}    
    #summarise data by day
    myIntervals<-aggregate(steps ~ interval, myData, FUN = mean, rm.na = TRUE)
    plot(myIntervals, type = "l")  
    text(240, 200, paste("Maximum interval = ", 
                         which(myIntervals$steps == max(myIntervals$steps))))
```

## Imputing missing values
As noted above there are a number of missing values which are coded as NA. The exact number is calculated by the code:
```{r}
    sum(is.na(myData$steps))
```
A good way to replace these so that they do not distort the data is by replacing them with the mean response for their interval:
```{r}
    # Revise data to replace missing values
    myNAs <- which(is.na(myData$steps))
    newData <- myData
    newData[myNAs, "steps"] <- myIntervals[newData[myNAs, "interval"], "steps"]
```

The following snapshots illustrate how this affects the data:

```{r}
    tail(myData, 5)

    tail(newData, 5)
```

As a result of these changes, the histogram prepared from this new data set has changed slightly: 
```{r}
    # create histogram of average steps per day with revised data
    newSteps <- aggregate(steps ~ date, newData, FUN = mean)
    hist(newSteps$steps, main = "Histogram of Steps per day", xlab = "")
    text(70, 18, paste("Mean = ", round(mean(newSteps$steps), 2), "\n Median = ", 
                     round(median(newSteps$steps), 2)))
    ```

It appears that replacing the NA's with the mean for their 5-minute interval has had only a minor effect upon the data since both the mean and median have not changed.  

## Are there differences in activity patterns between weekdays and weekends?
There are clear differences in activity patterns between weekdays and weekends.  This can be seen from the following plot:
```{r}
    #Add column to identify weekday/weekend
    newData$day<-"Weekday"
    newData$day[weekdays(newData$time) %in% c("Saturday", "Sunday")] <- "Weekend"

    #Summarise and plot data
    newIntervals <- aggregate(steps ~ day * interval, data = newData, FUN = mean)
    xyplot(steps ~ interval | day, newIntervals, layout = c(1, 2), type = "l")
```
    